Tree Models

### 2025-12-10 Data Cleaning.Rmd

## RANDOM FOREST
rf_spec <- rand_forest(mtry = sqrt(.cols())) |>
  set_engine("randomForest", importance = TRUE) |>
  set_mode("classification")

@@

## Notes
# 34% accuracy for all predictors
# 47% accuracy for just temperature and pH

>>>>

## RANDOM FOREST
rf_spec <- rand_forest(mtry = sqrt(.cols()), trees=500, min_n=1) |>
  set_engine("randomForest", importance = TRUE) |>
  set_mode("classification")

@@

## Notes
# 34% accuracy for all predictors
# 46.6% accuracy for just temperature and pH

### 2025-12-10 Presentation DSCI445.Rmd

Similar to the stock dataset we've seen in labs and homeworks, our time series dataset needs lagged variables.

@@

We also introduced a categorical predictor indicating whether the dose increased, decreased, or stayed the same compared to the day before.

@@

**GOAL: Interpretable and Captures Nonlinear Behavior**

@@

# Coagulant Dose Change Prediction: Modeling

#

@@

@@

>>>>

Similar to the stock dataset we've seen in labs and homeworks, our time series dataset needs **lagged variables**.

@@

We also introduced a **categorical predictor** indicating whether the dose increased, decreased, or stayed the same compared to the day before.

@@

**Priorities: Interpretability, Capturing of Nonlinear Behavior**

@@

# Coagulant Dose Change Prediction: Multinomial, LDA/QDA

# Coagulant Dose Change Prediction: Random Forest

@@

__ performed the best out of models tested.

@@

Overall ineffective, and prone to overfitting (especially random forest).  
Model predicting no dose change would be more accurate.

**Insert bar graph**

# Conclusion and Possible Next Steps

- Predicting chemical doses based on these raw water characteristics proved to be ineffective
- Most likely reasons:
   - Limited size of dataset
   - Limited predictor set
   - Operator-dependent
   - Complexity of chemical relationships
   - Other factors (example: reservoir turnover)
- Operators make decisions based on variety of other factors: chemical waste production, filter performance, lab tests, etc.
- Incorporate more varieties and time ranges of data in future modeling efforts

### 2025-12-10 Tree Models Reece C.Rmd

library(dplyr)
library(tidyverse)
library(tidymodels)
#library(yardstick)
library(vip)

@@

model_metrics = predictions %>%
  metrics(truth = FeCl_dose, estimate = .pred) # Change target variable here

print(model_metrics)

ggplot(predictions, aes(x = FeCl_dose, y = .pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  labs(title = "Random Forest Accuracy",
       x = "Actual",
       y = "Predicted") +
  theme_bw()

vip(fit_rf)

>>>>

library(dplyr)
library(tidyverse)
library(tidymodels)
library(vip)
library(knitr)
library(kableExtra)

@@

model_metrics = predictions %>%
  metrics(truth = FeCl_dose, estimate = .pred) # Change target variable here
rmse = model_metrics$.estimate[1]
r_squared = model_metrics$.estimate[2]
mae = model_metrics$.estimate[3]

kable(metrics, booktabs = TRUE, digits = 3) %>%
  kable_styling(position = "center", font_size = 12)
model_metrics$.estimate

#ggplot(predictions, aes(x = FeCl_dose, y = .pred)) +
#  geom_point() +
#  geom_abline(intercept = 0, slope = 1) +
#  labs(title = "Random Forest Accuracy",
#       x = "Actual",
#       y = "Predicted") +
#  theme_bw()

vip(fit_rf, num_features = 20) +
  geom_col(color = "black", fill = "coral") +
  labs(
    title = "Feature Importance (Random Forest)",
    x = "Importance",
    y = "Feature"
  ) +
  theme_minimal(base_size = 14)
```
